# Skin Transcriptome in Time and Space

## Data cleaning and curation
```{r load package, echo=FALSE, message=FALSE, warning=FALSE}
library(tibble)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(purrr)
library(tidybulk)
library(tidySummarizedExperiment)
library(ComplexHeatmap)
library(DT)
library(BiocParallel)
library(DESeq2)
knitr::opts_chunk$set(cache = TRUE, message = FALSE, warning = FALSE, echo = FALSE)
```

```{r load data, cache=TRUE}
# read raw counts ---------------------------------------
counts <-
  gzfile("../multiomics-ad-transcriptomics/data/counts/counts_gene_name_redo.txt.gz") %>%
  readr::read_tsv(skip = 1) %>%
  mutate(Chr = Chr %>% str_extract("chr\\w{1,}(?=;)|chr\\w{1,}$"), #keeps only 1
         Strand = Strand %>% str_extract("\\+|\\-"),               #keeps only 1
         Start = Start %>% str_extract("\\d{1,}(?=;)|\\d{1,}$"),   #keeps only 1
         End = End %>% str_extract("\\d{1,}(?=;)|\\d{1,}$")) %>%   #keeps only 1
  filter(!Geneid %>% 
           stringr::str_detect("\\d{1,}P$|\\d{1,}P\\d{1,}$|\\.|-AS\\d{1}|-DT")) %>% # apply gene filtering (pseudo genes, antisense genese, DT genes)
  dplyr::rename_with(str_extract, starts_with("/home/projects/"), pattern = "NG-[:graph:]{1,}$")

# lib to merge -------------
lib_id <- colnames(counts)[-1:-6] %>% str_extract("lib\\d{1,}") # extract lib id
lib_id_t <- table(lib_id)
lib_id_merge <- lib_id_t[(lib_id_t != 1)] %>% rownames() #lib_ids to be merged

# merge technical replicates -------------------------------------------
source("../multiomics-ad-transcriptomics/R/helper.R")

counttable_LibMerged <- 
  bind_cols(counts[, 1:6], 
            counttable_merge_library_fun(counts %>% select(-1:-6), lib_id_merge)) 

# clinical records ------------------------
clinical_records <- 
  readr::read_csv("../multiomics-ad-transcriptomics/data/LEO-AD-metadata.csv") %>% 
  mutate(id = id %>% str_replace("-", "_"),
         JOIN_ID = paste(id, paste0("0",visit_no), sep="_"))

extensive_meta <-
  readr::read_csv("../multiomics-ad-transcriptomics/data/RNAseq_sample_annotation(extensive).csv")
```

```{r SummarizedExperiment}

# se col_data
col_data <-
  tibble(
    # definingn metadata --------------------------------------
    BAM_ID = colnames(counttable_LibMerged)[-1:-6]) %>% 
    mutate(sequencing_batch = BAM_ID %>% str_extract("NG-\\d{1,}"),
           group = BAM_ID %>% str_extract("AD|CO") %>% 
               forcats::fct_relevel(c("CO", "AD")),
           subject = BAM_ID %>% str_extract("(AD|CO)_\\d{2,}"),
           visit = BAM_ID %>% str_extract("\\d{2,}_(AD|CO)") %>% str_extract("\\d{2,}"),
           skin_type = BAM_ID %>% str_replace("TS", "LS") %>% str_extract("NN|NL|LS") %>% forcats::fct_relevel(c("NN", "NL", "LS")),
           replicate_ID = BAM_ID %>% str_replace("TS_BI", "BI_LS") %>% str_extract("(NN|NL|LS)_\\d{2,}") %>% str_extract("\\d{2,}"),
           library_ID = BAM_ID %>% str_extract("lib\\d{1,}"),
           sequencing_id = BAM_ID %>% str_extract("\\d{4}_(1|2)"),
           JOIN_ID = base::paste(subject, visit, sep = "_")) %>%
    left_join(clinical_records %>% select(id, gender) %>% filter(!is.na(gender)), by=c("subject"="id")) %>%
    left_join(clinical_records %>% select(JOIN_ID, scorad, scorad_objective, easi_total_score, date_visit)) %>% 
    select(-JOIN_ID) %>% 
    mutate(gender = gender %>% forcats::fct_relevel(c("female", "male"))) %>% 
    left_join(extensive_meta %>% select(BAM_ID, rna_quality, visit_quarter))

# Summarized Experiment--------------------
se <- SummarizedExperiment::SummarizedExperiment(
  assays = list(counts = counttable_LibMerged %>% select(-1:-6) %>% as.matrix),
  colData = col_data
)
names(se) <- counttable_LibMerged$Geneid
```

```{r filtering}
# filter out genes 
se <- se %>% keep_abundant(factor_of_interest = skin_type) %>% scale_abundance()

# filter out samples ----------------
se <- se[, se$BAM_ID != (assay(se) %>% colSums() %>% which.min %>% names)] # 392
se <- se[, !se$library_ID %in% c("lib390174", "lib390716")] # 390
```

```{r revise some sp info}
index_ad03v01ls01 <-
  which(colData(se)$subject == "AD_03" &
        colData(se)$visit == "01" &
        colData(se)$skin_type == "LS" &
        colData(se)$replicate_ID == "01")
colData(se)$skin_type[index_ad03v01ls01] <- "NL"

index_ad03v01nl01 <-
  which(colData(se)$subject == "AD_03" &
          colData(se)$visit == "01" &
          colData(se)$skin_type == "NL" &
          colData(se)$replicate_ID == "01")
colData(se)$skin_type[index_ad03v01nl01] <- "LS"

index_ad13v03nl01 <-
  which(colData(se)$subject == "AD_13" &
        colData(se)$visit == "03" &
        colData(se)$skin_type == "NL" &
        colData(se)$replicate_ID == "01")
colData(se)$visit[index_ad13v03nl01] <- "05"
colData(se)$skin_type[index_ad13v03nl01] <- "LS"
colData(se)$replicate_ID[index_ad13v03nl01] <- "02"

index_ad13v05ls02 <-
  which(colData(se)$subject == "AD_13" &
        colData(se)$visit == "05" &
        colData(se)$skin_type == "LS" &
        colData(se)$replicate_ID == "02")
colData(se)$visit[index_ad13v05ls02] <- "03"
colData(se)$skin_type[index_ad13v05ls02] <- "NL"
colData(se)$replicate_ID[index_ad13v05ls02] <- "01"
```


```{r Sample QC}
library(googledrive)
library(googlesheets4)
gs4_auth(email = "tu.hu.ucph@gmail.com", token = drive_token())
sample_qc_record_url <- 
  googledrive::drive_ls("multiomics-ad-transcriptomics/") %>% 
  filter(name == "sample_qc_record") %>% 
  pull(id)
sample_qc_record_edited <-
  googlesheets4::read_sheet(ss = sample_qc_record_url)
```

```{r exclude}
exclude_BAM <- 
  colData(se) %>% as_tibble() %>% 
  select(BAM_ID, subject, visit, skin_type, replicate_ID) %>%
  left_join(sample_qc_record_edited) %>% 
  filter(exclude == TRUE) %>% pull(BAM_ID)
```

```{r generate new se, cache.lazy=FALSE}
se <- 
  se %>% 
  filter(!BAM_ID %in% exclude_BAM)
```
## Explorative data analysis
### Principle component analysis (PCA)
```{r PCA, include = TRUE}
PCA <- 
  se %>% 
  reduce_dimensions(method = "PCA", top = 350)

PCA %>% 
  pivot_sample() %>% 
  ggplot(aes(PC1, PC2, 
             color = skin_type, 
             scorad = scorad,
             subject = subject, visit = visit, 
             skin_type = skin_type, replicate_ID = replicate_ID,
             rna_quality = rna_quality)) + geom_point() +
  xlab(paste0("PC1 (Var% explained:", summary(attr(PCA, "internals")$PCA)$importance[2,1] %>% scales::percent(), ")")) +
  ylab(paste0("PC2 (Var% explained:", summary(attr(PCA, "internals")$PCA)$importance[2,2] %>% scales::percent(), ")")) +
  theme_classic() + 
  ggsci::scale_color_nejm() +
  ggtitle("PCA plot")
```

```{r cleaning}
rm(list = ls()[ls() != "se"])
```

### Sample Distance and Similarity
```{r distance}
dist_st <- # st: skin_type
  lapply(c("LS", "NL", "NN"), function(x){
    se %>% 
      subset(select = skin_type == x) %>% 
      assay(2) %>% 
      t() %>% 
      dist() %>% 
      as.numeric()})

cosine_st <-
    lapply(c("LS", "NL", "NN"), function(x){
    se %>% 
      subset(select = skin_type == x) %>% 
      assay(2) %>% 
      coop::cosine() %>% 
      as.numeric()})

names(dist_st) <- c("LS", "NL", "NN")
names(cosine_st) <- c("LS", "NL", "NN")

dist_df <- plyr::ldply(dist_st, tibble, .name_repair = ~ c("dist"),
                       .id = "skin_type")

cosine_df <- plyr::ldply(cosine_st, tibble, .name_repair = ~ c("cos_dist"),
                       .id = "skin_type")

dist_df %>% 
  ggplot(aes(x = dist, 
             color = skin_type,
             fill = skin_type)) +
  geom_density(alpha = .6) +
  theme_classic() +
  ggsci::scale_color_nejm() + 
  ggsci::scale_fill_nejm() +
  xlab("Euclidean Distance")

cosine_df %>% 
  ggplot(aes(x = cos_dist, 
             color = skin_type,
             fill = skin_type)) +
  geom_density(alpha = .4) +
  theme_classic() +
  ggsci::scale_color_nejm() + 
  ggsci::scale_fill_nejm() +
  xlab("Cosine similarity")

```

### Heatmap
```{r heatmap, fig.align='center', dpi = 300}
heatmap_data <- 
  se %>% 
  keep_variable(.abundance = counts_scale,
                top = 150)

assay(heatmap_data) <-
  log1p(assay(heatmap_data))

heatmap_data_m <- heatmap_data %>% assay()

heatmap_annotation <- 
  HeatmapAnnotation(`tissue type` = colData(heatmap_data) %>% as_tibble() %>% pull(skin_type))

ComplexHeatmap::Heatmap(heatmap_data_m, 
                        name = "log(gene expression)",
                        show_column_names = FALSE,
                        row_names_gp = gpar(fontsize = min(12, 400 / dim(heatmap_data_m)[1])),
                        top_annotation = heatmap_annotation)
```

## Differential gene expression (DGE) analysis

### Baseline

```{r DGE-V1, echo=TRUE}
# cleaning
rm(list = ls()[ls() != "se"])

register(MulticoreParam(7))
# V1 LSvsNL --------------------
se_v1_LSvsNL <- se[, se$visit == "01" & se$skin_type %in% c("LS", "NL")]
de_v1_LSvsNL <- DESeq2::DESeqDataSet(se_v1_LSvsNL, ~ subject + skin_type) %>% DESeq2::DESeq(parallel = TRUE)
de_v1_LSvsNL_LFC <- lfcShrink(de_v1_LSvsNL, coef = "skin_type_LS_vs_NL", type = "apeglm", parallel = TRUE, quiet = TRUE)
de_v1_LSvsNL_result <- bind_cols(tibble(gene_name = rownames(se_v1_LSvsNL)), de_v1_LSvsNL_LFC %>% as_tibble())
de_v1_LSvsNL_result_s <- de_v1_LSvsNL_result %>% filter(padj < .05, abs(log2FoldChange) > 1)
# V1 LSvsNN --------------------
se_v1_LSvsNN <- se[, se$visit == "01" & se$skin_type %in% c("LS", "NN")]
de_v1_LSvsNN <- DESeq2::DESeqDataSet(se_v1_LSvsNN, ~ gender + skin_type) %>% DESeq2::DESeq(parallel = TRUE)
de_v1_LSvsNN_LFC <- lfcShrink(de_v1_LSvsNN, coef = "skin_type_LS_vs_NN", type = "apeglm", parallel = TRUE, quiet = TRUE)
de_v1_LSvsNN_result <- bind_cols(tibble(gene_name = rownames(de_v1_LSvsNN)), de_v1_LSvsNN_LFC %>% as_tibble())
de_v1_LSvsNN_result_s <- de_v1_LSvsNN_result %>% filter(padj < .05, abs(log2FoldChange) > 1)
# V1 LSvsNN --------------------
se_v1_NLvsNN <- se[, se$visit == "01" & se$skin_type %in% c("NL", "NN")]
de_v1_NLvsNN <- DESeq2::DESeqDataSet(se_v1_NLvsNN, ~ gender + skin_type) %>% DESeq2::DESeq(parallel = TRUE)
de_v1_NLvsNN_LFC <- lfcShrink(de_v1_NLvsNN, coef = "skin_type_NL_vs_NN", type = "apeglm", parallel = TRUE, quiet = TRUE)
de_v1_NLvsNN_result <- bind_cols(tibble(gene_name = rownames(de_v1_NLvsNN)), de_v1_NLvsNN_LFC %>% as_tibble())
de_v1_NLvsNN_result_s <- de_v1_NLvsNN_result %>% filter(padj < .05, abs(log2FoldChange) > 1)
```

```{r Consolidate Results display}
de_v1_c <-
  bind_rows(de_v1_LSvsNL_result_s %>% mutate(contrast = "LSvsNL"),
            de_v1_LSvsNN_result_s %>% mutate(contrast = "LSvsNN"),
            de_v1_NLvsNN_result_s %>% mutate(contrast = "NLvsNN")) %>% 
  select(gene_name, log2FoldChange, padj, contrast) %>% 
  pivot_wider(names_from = contrast, values_from = c("log2FoldChange", "padj")) %>% 
  arrange(-log2FoldChange_LSvsNL)

t_container <- htmltools::withTags(table(
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'Gene name'),
      th(colspan = 3, 'log2 Fold Change'),
      th(colspan = 3, 'Adjusted p-value')
    ),
    tr(
      lapply(rep(c('LSvsNL', 'LSvsNN', "NLvsNN"), 2), th)
    )
  )
))
  
datatable(de_v1_c, colnames = c("gene name", "LSvsNL", "LSvsNN", "NLvsNN", "LSvsNL", "LSvsNN", "NLvsNN"), container = t_container, rownames = FALSE,
          options = list(pageLength = 20)) %>% 
  formatRound(de_v1_c %>% colnames %>% str_extract("log2FoldChange_\\w{6}") %>% purrr::discard(is.na)) %>% 
  formatSignif(de_v1_c %>% colnames %>% str_extract("padj_\\w{6}") %>% purrr::discard(is.na), digits = 3)
```

Number of dysregulated genes at visit 1:

* `LSvsNL`: `r de_v1_LSvsNL_result_s %>% filter(log2FoldChange > 0) %>% nrow` up-regulated and `r de_v1_LSvsNL_result_s %>% filter(log2FoldChange < 0) %>% nrow` down-regulated;
* `LSvsNN`: `r de_v1_LSvsNN_result_s %>% filter(log2FoldChange > 0) %>% nrow` up-regulated and `r de_v1_LSvsNN_result_s %>% filter(log2FoldChange > 0) %>% nrow` down-regulated
* `NLvsNN`: `r de_v1_NLvsNN_result_s %>% filter(log2FoldChange > 0) %>% nrow` up-regulated and `r de_v1_NLvsNN_result_s %>% filter(log2FoldChange < 0) %>% nrow` down-regulated


### Time Effect

I performed `Likelihood Ratio Test (LRT)` to test the genes react in the condition-specific manner over time, compared to baseline samples. I first specify a `full model` containing all terms. Then test a `reduced model`.

```{r model time (visit)}

# clean the environment for parallel 
rm(list = ls()[ls() != "se"])

# LS vs NL
se_t_LSvsNL <- se %>% filter(skin_type %in% c("LS", "NL"))
de_t_LSvsNL <- DESeq2::DESeqDataSet(se_t_LSvsNL, ~ subject + visit + skin_type + skin_type * visit)

de_t_LSvsNL_m <- DESeq2::DESeq(de_t_LSvsNL, test="LRT", 
                              reduced = ~  subject + visit + skin_type,
                              parallel = TRUE)

deT_LSvsNL_res <- 
  bind_cols(tibble(gene_name = names(de_t_LSvsNL_m)),
  DESeq2::results(de_t_LSvsNL_m) %>% as_tibble())

# LS vs NN
se_t_LSvsNN <-
  se %>%
  filter(skin_type %in% c("LS", "NN"))

de_t_LSvsNN <- DESeq2::DESeqDataSet(se_t_LSvsNN, ~  gender + visit + skin_type + skin_type * visit)

de_t_LSvsNN_m <- DESeq2::DESeq(de_t_LSvsNN, test="LRT", 
                              reduced = ~ gender + visit + skin_type,
                              parallel = TRUE)
de_t_LSvsNN_res <- 
  bind_cols(tibble(gene_name = names(de_t_LSvsNN_m)),
  DESeq2::results(de_t_LSvsNN_m) %>% as_tibble())

# NL vs NN
se_t_NLvsNN <-
  se %>%
  filter(skin_type %in% c("NL", "NN"))

de_t_NLvsNN <- DESeq2::DESeqDataSet(se_t_NLvsNN, ~  gender + visit + skin_type + skin_type * visit)

de_t_NLvsNN_m <- DESeq2::DESeq(de_t_NLvsNN, test="LRT", 
                              reduced = ~ gender + visit + skin_type,
                              parallel = TRUE)
de_t_NLvsNN_res <- 
  bind_cols(tibble(gene_name = names(de_t_NLvsNN_m)),
  DESeq2::results(de_t_NLvsNN_m) %>% as_tibble())
```

```{r model time(visit)-vis}
betas <- coef(de_t_LSvsNN_m)
colnames(betas)
topGenes <- names(de_t_LSvsNN_m)[head(order(de_t_LSvsNN_res$padj),20)]
mat <- betas[topGenes, -c(1,2)]
thr <- 3 
mat[mat < -thr] <- -thr
mat[mat > thr] <- thr
pheatmap(mat, breaks=seq(from=-thr, to=thr, length=101),
         cluster_col=FALSE)
```



```{r model time(quarter)}

se <- se[, - which(is.na(se$visit_quarter))]
se_q_LSvsNL <- se[, se$skin_type %in% c("LS", "NL")]
se_q_LSvsNN <- se[, se$skin_type %in% c("LS", "NN")]
se_q_NLvsNN <- se[, se$skin_type %in% c("NL", "NN")]

# LS vs NL

de_q_LSvsNL <- DESeq2::DESeqDataSet(se_q_LSvsNL, ~ subject + visit_quarter + skin_type + skin_type * visit_quarter)

de_q_LSvsNL_m <- DESeq2::DESeq(de_q_LSvsNL, test="LRT", 
                              reduced = ~  subject + visit_quarter + skin_type,
                              parallel = TRUE)

de_q_LSvsNL_res <- 
  bind_cols(tibble(gene_name = names(de_q_LSvsNL_m)),
  DESeq2::results(de_q_LSvsNL_m) %>% as_tibble())

# LS vs NN

de_q_LSvsNN <- DESeq2::DESeqDataSet(se_q_LSvsNN, ~  gender + visit_quarter + skin_type + skin_type * visit_quarter)

de_q_LSvsNN_m <- DESeq2::DESeq(de_q_LSvsNN, test="LRT", 
                              reduced = ~ gender + visit_quarter + skin_type,
                              parallel = TRUE)
de_q_LSvsNN_res <- 
  bind_cols(tibble(gene_name = names(de_q_LSvsNN_m)),
  DESeq2::results(de_q_LSvsNN_m) %>% as_tibble())

# NL vs NN
de_q_NLvsNN <- DESeq2::DESeqDataSet(se_q_NLvsNN, ~  gender + visit_quarter + skin_type + skin_type * visit_quarter)

de_q_NLvsNN_m <- DESeq2::DESeq(de_q_NLvsNN, test="LRT", 
                              reduced = ~ gender + visit_quarter + skin_type,
                              parallel = TRUE)
de_q_NLvsNN_res <- 
  bind_cols(tibble(gene_name = names(de_q_NLvsNN_m)),
  DESeq2::results(de_q_NLvsNN_m) %>% as_tibble())
```

```{r model time(quarter)-vis}
heatmap_de_model <- 
  function(de_model){
    de_model_res <- 
      bind_cols(tibble(gene_name = names(de_model)),
                DESeq2::results(de_model) %>% as_tibble())
    betas <- coef(de_model)
    
    topGenes <- names(de_model)[head(order(de_model_res$padj),20)]
    col_to_plot <- colnames(betas) %>%
                   str_extract("visit_quarter[:graph:]{1,}") %>% 
                   purrr::discard(is.na)
    mat <- betas[topGenes, col_to_plot]
    colnames(mat) <- col_to_plot %>% str_remove("visit_quarter") %>% str_remove("^_")
    thr <- 3 
    mat[mat < -thr] <- -thr
    mat[mat > thr] <- thr
    pheatmap(mat, 
             breaks=seq(from=-thr, to=thr, length=101), 
             cluster_col=FALSE, angle_col = "45", 
             main = deparse(substitute(de_model)) %>% str_extract("(LS|NL)vs(NL|NN)"))
  }

heatmap_de_model(de_q_LSvsNL_m)
heatmap_de_model(de_q_LSvsNN_m)
heatmap_de_model(de_q_NLvsNN_m)
```


### Space effect


## Gene Set Analysis

## Cell Type Deconvolution
```{r}

```


## Varian Partition (to work)

```{r input data, eval=FALSE, echo=TRUE}

# metadata -------------------------
metadata <- 

# -----------------------------------

# gene expression data -------------
countable <- dl$countable[,metadata$BAM_ID] %>% as.matrix()
rownames(countable) <- dl$gene_annotation$Geneid
geneExpr <- DGEList(countable)
geneExpr <- calcNormFactors(geneExpr)

# formula --------------------------
countable <- dl$countable[,metadata$BAM_ID] %>% as.matrix()
rownames(countable) <- dl$gene_annotation$Geneid
geneExpr <- DGEList(countable)
geneExpr <- calcNormFactors(geneExpr)
form <- ~ skin_type + (1|subject) + visit
v_obj_dream <- voomWithDreamWeights(geneExpr, form, metadata)
```


```{r variance explained, eval=FALSE, echo=TRUE}

vp <- 
  variancePartition::fitExtractVarPartModel(
    assay(se, 2), 
    ~ (1|skin_type) + (1|subject) + (1|visit) + (1|gender), 
    colData(se) %>% as_tibble()
    )
variancePartition::plotVarPart(vp)
```

```{r}
se_500 <- se %>% keep_variable()
vp_500 <- variancePartition::fitExtractVarPartModel(
    assay(se_500, 2), 
    ~ (1|skin_type) + (1|subject) + (1|visit) + (1|gender), 
    colData(se_500) %>% as_tibble()
    )
variancePartition::plotVarPart(vp_500)
```

